---
title: "Thoughts on Plots"
subtitle: "Real examples of data analytics and visualization expertise"
---

::: {.page-header}
Explore interactive visualizations, analytical approaches, and techniques that demonstrate my capabilities in transforming data into insights.
:::

## Featured Visualizations {#visualizations}

::: {.demo-grid}

### Advanced ggplot2 Techniques

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 6

# Your existing R visualization code here
library(ggplot2)
library(dplyr)

# Example plot - replace with your actual demos
```

**Key Features:**
- Custom themes and styling
- Complex data transformations
- Interactive elements with plotly

---

### Time Series Analysis

```{r}
#| echo: true

# Your time series visualization code
```

**Demonstrates:**
- Trend analysis
- Seasonal decomposition
- Forecasting methods

---

### Interactive Dashboards

```{r}
#| echo: true

# Your interactive visualization code
```

**Technologies:**
- R Shiny
- Plotly
- Custom HTML widgets

:::

## Case Studies

::: {.case-study}
### Project: Market Analysis Dashboard

**Challenge:** Client needed to understand customer behavior patterns across multiple channels.

**Solution:** Built comprehensive analytical framework with interactive visualizations.

**Results:**
- 40% improvement in insight discovery time
- Identification of 3 new market segments
- Improved decision-making process

[View Full Case Study →](#)
:::

::: {.case-study}
### Project: Predictive Modeling Pipeline

**Challenge:** Automate forecasting process for inventory management.

**Approach:** Developed custom R package with modeling and visualization components.

**Outcome:**
- 25% reduction in forecast error
- Streamlined workflow
- Reproducible reporting system

[View Technical Details →](#)
:::

---

## Additional Examples

Browse more demonstrations organized by technique and application:

::: {.example-links}
- [Statistical Modeling Examples →](#)
- [Data Cleaning & Transformation →](#)
- [Geospatial Visualizations →](#)
- [Network Analysis →](#)
- [Text Analytics →](#)
:::

---

[← Back to Home](index.qmd) | [View Apps →](apps.qmd)